Neural Models for Target-Based Computer-Assisted Musical Orchestration: A Preliminary Study

### 2021-06-28 csmc_mume_2020/paper/JCMS_2021/paper.tex
Add revisions from JCMS reviews

The approach studied in \cite{Carpentier2010} consists in finding a good orchestration for any given sound by searching combinations of sounds from a database with a multi-objective optimization heuristics and a constraint solver that are jointly optimized. Both the target sound and the sounds in the database are embedded in a feature space defined by a fixed feature function and each generated combination of sounds is evaluated by using a specific metric. This method has been substantially improved in \cite{Cella18, Cella2020} and is implemented in the \emph{Orchidea} toolbox for assisted orchestration (\url{www.orch-idea.org}), currently considered the state-of-the-art system for assisted orchestration.

In this paper, we try a different approach to this problem by experimenting with deep neural architectures. Unlike Orchidea, which uses an evolutionary optimization algorithm to find a solution, we frame computer-assisted orchestration as a classification task. The main idea is to train a model to classify combinations of real instruments and then use this model for orchestration. Where Orchidea has multiple pre- and post-processing steps, we create an end-to-end network that takes a target sound as input and outputs a list of samples that can be combined to create the orchestrated solution. A typical solution for assisted orchestration is a set of triples \emph{instrument-pitch-dynamics} such as \{\texttt{Flute C6 pp, Bassoon C4 mf, Bassoon G4 ff}\}. By training a neural network with real combinations of instrumental notes, it will acquire the ability to identify the presence of each instrument and its associated pitch by building the appropriate latent representation. Thus, when an unknown target sound is given as input, the network will identify which are the best instruments to match the target sound, and it will be able to  deconstruct a complex mixture of timbres into individual instrument notes. This method is motivated by the good results obtained in previous research on musical instruments identification \cite{Benetos07, Kitahara05} and the more recent use of deep neural networks for musical classification \cite{lostanlen16, Bian19}.

We perform preliminary experiments with two deep architectures: a convolutional neural network (CNN) with a long short-term memory (LSTM) unit in the middle and \emph{ResNet}, a well known residual architecture that already yielded good results for image classification \cite{He15}. We choose to use a CNN because of its success in audio classification \cite{Hershey17} and include an LSTM unit in it because of its ability to learn long term dependencies in data \cite{Hochreiter97}, which is important given the temporal nature of audio.

@@

\subsection{From Orchestration to Classification}

@@

The choice of using the Mel spectrogram as input features for classification models is common in music information retrieval \cite{McKinney2003} and can be considered to be an appropriate representation of sound and musical signals. Such setup proved to be successful in \cite{Salamon17}.

@@

In order to get a better sense of the complexity of the problem, we tested three baseline classifiers: support vector machine (SVM), random forest (RF), and K-nearest neighbors (KNN). We used the implementations provided in the \emph{scikit-learn }library for Python \cite{scikit-learn}. More specifically, for SVM we used a non-linear RBF kernel and for RF we set the maximum depth of each tree to be 15.  All of the baseline experiments used 50,000 generated samples with a train-test split of 60/40. Each sample was made by a combination of one to ten instruments and was four seconds in length.

@@

The complete problem of classification of the pairs instrument-pitch has 424 total classes. This problem is very difficult for parametric classifiers, so we initially switched to the simpler problem of classifying only the instruments and not the pitch. This had the benefit of both reducing the number of classes and increasing the number of samples per class. We found that SVM was able to very accurately identify the instrument given an input that had only one instrument present; for this case the accuracy was 99.8\%. However, as soon as multiple instruments were present in the input, the accuracy dropped significantly. With two instruments, accuracy was 55.4\%, with three it was 19.6\%. KNN performed significantly worse than SVM, so we did not attempt any further testing with it.

To better approximate our original problem of identifying instrument and pitch, however, we performed experiments in which two instruments were selected and for input data that contained samples from one of those two instruments, both the instrument and pitch \emph{class} of the sample would be classified. The pitch class is the note name without the octave, e.g. C, D\#, G, etc. The input was a combination of two instruments drawn from a possible twelve instruments, and the classifier attempted to identify which instruments were present and for the specified instruments, say Violin and Viola, which pitch classes were present. If another instrument was present in the input combination that was not Violin or Viola, the classifier would simply identify that an instrument that was not one of the two was present. The accuracies from this experiment are outlined in Table \ref{tab:baselines}. Since RF performed worse than SVM in every experiment, we stopped testing with it and used SVM from that point on.

@@

This was still a simplified version of the problem, as we were identifying only the pitch class of a few instruments. However, the parametric classifiers were unable to achieve accurate results as the number of instruments increased. Therefore, we did not attempt, with parametric classifiers, the full setting of the problem in which individual pitches are classified for all instruments, and instead turned to our neural models.

@@

The CNN was trained on 200,000 generated samples over 50 epochs, and ResNet was trained on 400,000 generated samples over 20 epochs. Fig. \ref{cnn_vs_resnet} shows the best overall test accuracy achieved by both models across the number of samples $N$.
ResNet outperforms the CNN regardless of the number of samples used in the combination. This result is consistent with previous research \cite{He15}, as residual networks usually perform well in classification problems.

@@

While our models are not able to outperform Orchidea, they shows consistent results. During training, CNN and ResNet both perform well, with ResNet consistently achieving higher training accuracies than CNN. For orchestration, CNN performs quantitatively better with an average ratio of $0.71$ compared to ResNet with an average ratio of $0.66$ (see Table \ref{orch_eval}). Both models outperform Orchidea on three out of fifteen of the targets. Through our qualitative inspection, we find that CNN seems to better emulate the timbre in its orchestrations, where ResNet is better for recreating the harmonic content of the target.

>>>>

The approach studied in \cite{Carpentier2010} consists in finding an orchestration for any given sound by searching combinations of sounds from a database with a multi-objective optimization heuristic and a constraint solver that are jointly optimized. Both the target sound and the sounds in the database are embedded in a feature space defined by a fixed feature function and each generated combination of sounds is evaluated by using a specific metric. This method has been substantially improved in \cite{Cella18, Cella2020} and is implemented in the \emph{Orchidea} toolbox for assisted orchestration (\url{www.orch-idea.org}), currently considered the state-of-the-art system for assisted orchestration.

In this paper, we try a different approach to this problem by framing computer-assisted orchestration as a classification task. The main idea is to train a model to classify combinations of real instruments and then use this model for orchestration. To solve this classification problem, we turn to deep neural networks, which have shown success among a variety of classification tasks. Where Orchidea has multiple pre- and post-processing steps, we create an end-to-end network that takes a target sound as input and outputs a list of samples that can be combined to create the orchestrated solution.

A typical solution for assisted orchestration is a set of triples \emph{instrument-pitch-dynamics} such as \{\texttt{Flute C6 pp, Bassoon C4 mf, Bassoon G4 ff}\}. By training a neural network with real combinations of instrumental notes, it will acquire the ability to identify the presence of each instrument and its associated pitch by building the appropriate latent representation. Thus, when an unknown target sound is given as input, the network will identify which are the best instruments to match the target sound, and it will be able to  deconstruct a complex mixture of timbres into individual instrument notes. This method is motivated by the good results obtained in previous research on musical instruments identification \cite{Benetos07, Kitahara05} and the more recent use of deep neural networks for musical classification \cite{lostanlen16, Bian19}.

We perform preliminary experiments with two deep architectures: a convolutional neural network (CNN) with a long short-term memory (LSTM) unit in the middle and \emph{ResNet}, a well known residual architecture that has yielded positive results for image classification \cite{He15}. We choose to use a CNN because of its success in audio classification \cite{Hershey17} and include an LSTM unit in it because of its ability to learn long term dependencies in data \cite{Hochreiter97}, which is important given the temporal nature of audio.

@@

\subsection{From Classification to Orchestration}

@@

The choice of using the Mel spectrogram as input features for classification models is common in music information retrieval \cite{McKinney2003} and can be considered to be an appropriate representation of sound and musical signals. Such setup proved to be successful in \cite{Salamon17}. Specifically, the Mel spectrogram uses the Mel scale to create a perceptual representation of sound. Since we are attempting to recreate the timbre of a sound, and timbre is a perceptual descriptor, the Mel spectrogram is a fitting representation.

@@

In order to get a better sense of the complexity of the problem, we tested three baseline classifiers: support vector machine (SVM), random forest (RF), and K-nearest neighbors (KNN). We used the implementations provided in the \emph{scikit-learn }library for Python \cite{scikit-learn}. More specifically, for SVM we used a non-linear RBF kernel and for RF we set the maximum depth of each tree to be 15. All of the baseline experiments used 50,000 generated samples with a train-test split of 60/40. Each sample was made by a combination of one to ten instruments and was four seconds in length. At the beginning of each experiment, a new set of combinations was created.

@@

The complete problem of classification of the instrument-pitch pairs has 424 total classes. This problem is very difficult for parametric classifiers, so we initially switched to the simpler problem of classifying only the instruments and not the pitch. This had the benefit of both reducing the number of classes and increasing the number of samples per class. We found that SVM was able to very accurately identify the instrument given an input that had only one instrument present; for this case the accuracy was 99.8\%. However, as soon as multiple instruments were present in the input, the accuracy dropped significantly. With two instruments, accuracy was 55.4\%, with three it was 19.6\%. KNN performed significantly worse than SVM, so we did not attempt any further testing with it.

To better approximate our original problem of identifying instrument and pitch, however, we performed experiments in which two instruments were selected and for input data that contained samples from one of those two instruments, both the instrument and pitch \emph{class} of the sample would be classified. The pitch class is the note name without the octave, e.g. C, D\#, G, etc. The input was a combination of two instruments drawn from a possible twelve instruments, and the classifier attempted to identify which instruments were present and for the specified instruments, say Violin and Viola, which pitch classes were present. If another instrument was present in the input combination that was not Violin or Viola, the classifier would simply identify that an instrument that was not one of the two was present. A selection of the results from this experiment are outlined in Table \ref{tab:baselines}. Since RF performed worse than SVM in every experiment, we stopped testing with it and used SVM from that point on.

@@

This was still a simplified version of the problem, as we were identifying only the pitch class of a few instruments. However, the parametric classifiers were unable to achieve accurate results as the number of instruments increased. Therefore, we did not attempt, with parametric classifiers, the full setting of the problem in which individual pitches are classified for all instruments, and instead turned to neural models.

@@

The CNN was trained on 200,000 generated samples over 50 epochs, and ResNet was trained on 400,000 generated samples over 20 epochs. ResNet, being a deeper model than the CNN, requires more data per epoch during training. Fig.~\ref{cnn_vs_resnet} shows the best overall test accuracy achieved by both models across the number of samples $N$.
ResNet outperforms the CNN regardless of the number of samples used in the combination. This result is consistent with previous research \cite{He15}, as residual networks usually perform well in classification problems.

@@

While our models are not able to outperform Orchidea, they show consistent results. During training, CNN and ResNet both perform well, with ResNet achieving higher training accuracies than CNN. For orchestration, CNN performs quantitatively better with an average ratio of $0.71$ compared to ResNet with an average ratio of $0.66$ (see Table \ref{orch_eval}). Both models outperform Orchidea on three out of fifteen of the targets. Through our qualitative inspection, we find that CNN seems to better emulate the timbre in its orchestrations, where ResNet is better for recreating the harmonic content of the target.
